Contracts for unions and intersections are hard for Nickel

### 2021-06-06 paper.tex
Remove the \newpage-s
I missed a newpage
Small fixes

Union types, meaning a type \nickel{A @| B} containing values which
belong either to a type \nickel{A} or \nickel{B}, are a popular tool
when adding a static types to a dynamic language. In particular, both
TypeScript~\cite{TypeScriptUnions} and MyPy~\cite{MyPyOptional}, use
union typed to model the frequent practice to use the value
\lstinline{null} (\lstinline{None} in Python) to represent an absent
optional value. This is why the gradual typing literature, concerned
with formalising the interplay between static and dynamic type
systems, has been quite interested in union
types~\cite{RootCauseOfBlame,KeilThiemannUnionIntersection}.

@@

However, with the advent of DevOps, configurations have been extended
to describe the entire state of a computer, or even a fleet of
computers. For instance, with Kubernetes you need to configure a large
fleet of (possibly replicated) docker containers. To describe this
sort of configurations, you really want to be able to re-use and
abstract parts of the configuration, like traditional programming
languages do. To meet this need, languages such as Cue~\cite{?},
Dhall~\cite{?}, Jsonnet~\cite{?}, or Nickel~\cite{NickelRepo}, where
configurations are generated rather than spelt out, were created.

@@

These are inherently dynamic questions, as they are all questions
about the evaluated configuration. To this effect, Nickel lets us
annotate any expression with a dynamic schema check: \nickel{exp |
  C}. There is also syntactic sugar to annotate definitions:
\nickel{let id | C = value in exp} stands for \nickel{let id = (value
  | C) in exp}.\unsure{This point used to be made, to we want to
  reinsert it? However, Nickel programs are of a special kind: they are usually
simple, terminating programs that run on fixed inputs. Correctness errors will
show up at evaluation anyway, or will not matter (dead code).}

@@

Like in the case of lazy dictionaries, the checks are delayed.
Contracts keeps track of whether the caller or the implementation is
at fault for a violation, hence it can report precise error
messages. Compare Figure~\ref{fig:contract-reporting-wo} and
Figure~\ref{fig:contract-reporting-w}: in
Figure~\ref{fig:contract-reporting-wo} an error is reported inside the
\nickel{catHosts} function, but \nickel{catHosts} is, in fact,
correct, as is made clear by Figure~\ref{fig:contract-reporting-wo},
where \nickel{catHosts} is decorated with the \nickel{Str -> Str}
contract, and correctly reports that the caller failed to call
\nickel{catHosts} with a string argument.

@@

\newpage

@@

\newpage

\section{Union \& intersection}
\label{sec:union-inter}
\info{What are they, what are they used for.}
\resolved{Since nickel uses | for contract application, I'd move unions
to some other symbol.
I think the best option is to move to $\cap / \cup$}

@@

\change{These next few labels need to be inside figures,
they are not working}
\label{fig:union-adt-ex}
\begin{lstlisting}[title={Tree-like data structure}]
type BinaryTree =
Integer @| (BinaryTree, BinaryTree)
\end{lstlisting}

In addition to internal program representations, unions are also useful to model
external data:
\begin{itemize}
    \item To represent nullable values (\ref{fig:union-nullable-ex}).
    \item For plain enumerations (\ref{fig:union-enum-ex}).
    \item To support different representations of the same data
        (\ref{fig:union-alt-ex}).
\end{itemize}
All of which are common in configurations.

\label{fig:union-nullable-ex}
\begin{lstlisting}[title={Nullable values}]
optionalValue : Str @| Null
\end{lstlisting}

\label{fig:union-enum-ex}
\begin{lstlisting}[title={Plain enumeration}]
protocol : `Http @| `Ftp @| `Sftp
\end{lstlisting}

\label{fig:union-alt-ex}
\begin{lstlisting}[title={Alternative representations}]
person : {name : Str, age : Int}
  @| {name : Str, dateOfBirth : Str}
\end{lstlisting}

@@

This is only the beginning of the story. A language designer then has to choose
what contract combinators to provide among a number of possibilities: there is
the standard package including higher-order function contracts, record contracts
and list contracts.  Then, one could add polymorphic
contacts~\cite{BlameForAll}, dependent function
contracts~\ref{DependentContracts}, and so on. For validation purpose, union and
intersection contracts are really appealing. Unions have a intuitive
interpretation as modeling alternatives and thus appear implicitly or explicitly
in a lot of data structures.

>>>>

Union types, meaning a type \nickel{A @| B} containing values which
belong either to a type \nickel{A} or \nickel{B}, are a popular tool
when adding a static types to a dynamic language. In particular, both
TypeScript~\cite{TypeScriptUnions} and MyPy~\cite{MyPyOptional}, use
union types to model the frequent practice to use the value
\lstinline{null} (\lstinline{None} in Python) to represent an absent
optional value. This is why the gradual typing literature, concerned
with formalising the interplay between static and dynamic type
systems, has been quite interested in union
types~\cite{RootCauseOfBlame,KeilThiemannUnionIntersection}.

@@

However, with the advent of DevOps, configurations have been extended
to describe the entire state of a computer, or even a fleet of
computers. For instance, with Kubernetes you need to configure a large
fleet of (possibly replicated) docker containers. To describe this
sort of configurations, you really want to be able to re-use and
abstract parts of the configuration, like traditional programming
languages do. To meet this need, languages such as Cue~\cite{cueLang},
Dhall~\cite{dhallLang}, Jsonnet~\cite{jsonnetLang}, or Nickel~\cite{NickelRepo}, where
configurations are generated rather than spelt out, were created.

@@

These are inherently dynamic questions, as they are all questions
about the evaluated configuration. To this effect, Nickel lets us
annotate any expression with a dynamic schema check: \nickel{exp |
  C}. There is also syntactic sugar to annotate definitions:
\nickel{let id | C = value in exp} stands for \nickel{let id = (value
  | C) in exp}.\unsure{This point used to be made, do we want to
  reinsert it? However, Nickel programs are of a special kind: they are usually
simple, terminating programs that run on fixed inputs. Correctness errors will
show up at evaluation anyway, or will not matter (dead code).}

@@

Like in the case of lazy dictionaries, the checks are delayed.
Contracts keeps track of whether the caller or the implementation is
at fault for a violation, hence it can report precise error
messages. Compare Figure~\ref{fig:contract-reporting-wo} and
Figure~\ref{fig:contract-reporting-w}: in
Figure~\ref{fig:contract-reporting-wo} an error is reported inside the
\nickel{catHosts} function, but \nickel{catHosts} is, in fact,
correct, as is made clear by Figure~\ref{fig:contract-reporting-w},
where \nickel{catHosts} is decorated with the \nickel{Str -> Str}
contract, and correctly reports that the caller failed to call
\nickel{catHosts} with a string argument.

@@

@@

\section{Union \& intersection}
\label{sec:union-inter}
\info{What are they, what are they used for.}

@@

\begin{figure}[h]
\begin{lstlisting}[title={Tree-like data structure}]
type BinaryTree =
Integer @| (BinaryTree, BinaryTree)
\end{lstlisting}
\caption{A binary tree data type using unions}
\label{fig:union-adt-ex}
\end{figure}

In addition to internal program representations, unions are also useful to model
external data (\ref{fig:unions-use-cases}):
\begin{itemize}
    \item To represent nullable values.
    \item For plain enumerations.
    \item To support different representations of the same data.
\end{itemize}
All of which are common in configurations.

\begin{figure}[h]
% \label{fig:union-nullable-ex}
\begin{lstlisting}[title={Nullable values}]
optionalValue : Str @| Null
\end{lstlisting}

% \label{fig:union-enum-ex}
\begin{lstlisting}[title={Plain enumeration}]
protocol : `Http | `Ftp | `Sftp
\end{lstlisting}

% \label{fig:union-alt-ex}
\begin{lstlisting}[title={Alternative representations}]
person : {name : Str, age : Int}
  @| {name : Str, dateOfBirth : Str}
\end{lstlisting}
\caption{Various use cases of unions}
\label{fig:unions-use-cases}
\end{figure}

@@

This is only the beginning of the story. A language designer then has to choose
what contract combinators to provide among a number of possibilities: there is
the standard package including higher-order function contracts, record contracts
and list contracts.  Then, one could add polymorphic
contacts~\cite{BlameForAll}, dependent function
contracts~\cite{DependentContracts}, and so on. For validation purpose, union and
intersection contracts are really appealing. Unions have a intuitive
interpretation as modeling alternatives and thus appear implicitly or explicitly
in a lot of data structures.
